---
title: "Getting Started with COYU R package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with COYU R package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


# Introduction

The combined-over-years uniformity (COYU) criterion is used to assess
uniformity for measured, quantitative characteristics in DUS
trials. Uniformity is one of the criteria to be assessed before
granting protection of a new plant variety under the 
[UPOV system](https://www.upov.int).

COYU is used for measured characteristics, assessed over trials over
two or three years. The between-plant variability for a new variety is
compared to existing varieties planted in the same trials. To take
into account any relationship between level of expression and
variability, a spline- based adjustment is used. An older version of
this criterion used moving-averages rather than splines. The method is
described in UPOV document TGP/8 (version 5 and later) ["Trial Design
and Techniques Used in the Examination of Distinctness, Uniformity and
Stability"](https://www.upov.int/edocs/tgpdocs/en/tgp_8.pdf)

This package allows implementation of COYU. It requires variety means
in each trial for two or three years, along with pooled within-plot
standard deviations. It is freely available from the authors at
Biomathematics & Statistics Scotland (BioSS), via CRAN and GitHub (see
below).

# Installation

The R package has been written for R version 4 and above. If it is
running in an older version, it may not run. Successful application of
the package will require a reasonable level of R experience.

The package may be installed from CRAN using 

   ```{r, eval=FALSE}
   install.packages("coyu")
   ```

## Development

Firstly, the package `devtools' should be installed from CRAN. Then run the following
code

```{r, eval=FALSE}
library(devtools)
install_github("BiomathematicsAndStatisticsScotland/coyus/package/coyu")
```

The second command installs the current version of the COYU package
from GitHub.

The package does require the installation of another package lme4 and its dependencies,
which are available from CRAN. So please ensure that lme4 is installed prior to installing the
COYU package. We may investigate automating this at a later stage.

# Initial use

For every session, the R package should be called using:

```{r, eval=FALSE}
library(coyu)
```

There is help for each of the key functions along with a vignette
"GettingStarted". Note that the vignettes for the package may listed
using the command 

```{r, eval=FALSE}
vignettes(package='coyu')
```

# COYU package functions

The coyu package contains a number of functions designed to help produce the required
format for input and to structure output as well as running the COYU with spline analysis.
There are also a number of background functions that you will not need to access directly.
Here is a list of the key functions along with their use:

**COYU_all_results**

Core­function: carries the COYU analysis with spline adjustment and
generates a results set from trial data, parameters and probability
set objects.


**COYU_data_skeleton**

Input data manipulation: creates a "skeleton" data-frame of the right
size and type.

**COYU_parameters**

Input data manipulation: creates a parameter object with the correct
type. Usually called by COYU_parameters_from_df.

**COYU_parameters_from_df**

Input data manipulation: Infer parameters from the trial data
data-frame. You just need to supply the candidate AFP numbers.

**COYU_probability_set**

Input data manipulation: generates a probability set object from the
lists of probability values for 2 and 3 year rejection decisions along
with 2 year acceptance decisions.

**COYU_sanity_check**

Input data manipulation: checks consistency of input data

**COYU_print_results**

Output formatting: prints summary results following a run of
COYU_all_results

**COYU_results_as_dataframe**

Output formatting: converts output from COYU_all_results to a
convenient data-frame format. This can then be written to a file,
e.g. comma-separated value format using write.csv (csv files can
easily be read by Excel).

**COYU_plot_results**

Output formatting: produces plots for each character and year of the
spline curves fitted by COYU_all_results

# Running COYU

## Specification of input

In the first instance, we recommend that you use the function
COYU_all_results to run the COYU analyses using spline
adjustment. This function requires three arguments to define the input
for the analysis and it is suggested that you use the constructor
functions to create them:


### trial_data

trial data in a data ­frame with columns year, AFP, variety, one
column for each character (UP prefix followed by numeric identifier of
character) containing mean values, and one column for each character
(sUP prefix followed by numeric identifier of character) containing
within-plot standard deviations (mean over plots). Year is numeric
e.g. 2001. AFP denotes the numeric unique identifier for the varieties
and must be a factor. Variety is an optional text label for each
variety. Missing values are denoted NA as usual in R. This object
should have the S3 class "COYUs9TrialData"

### coyu_parameters 

a list defines which elements of the trial data are to be included in
the analysis. It is a list with elements candidates (a numeric vector
specifying AFP numbers of candidates), references (a numeric vector
specifying AFP numbers of reference varieties), characters (a numeric
vector specifying numbers of characters) and num_trial_years (a number
stating the number of years). This object should have the S3 class
"COYUs9Parameters".

### probability_sets

a two-dimensional array with each row defining a set of probability
values for decisions. The ability to define more than one set may be
useful in this practical exercise. There are three columns:

1. 3_year_reject (probability value for rejection after three years),
1. 2_year_reject (probability value for rejection after two years) and
2. 2_year_accept (probability value for acceptance after two years ­ this

will be a higher value than 2_year_reject).  Note that, with a 2 year
DUS test, 3_year_reject and 2_year_accept are redundant but the
software still requires values for these. This object should have the
S3 class "COYUs9ProbabilitySet".

## Argument constructor functions

To facilitate setting up of these files we have included utility functions
COYU_data_skeleton, COYU_parameters, COYU_parameters_from_df and
COYU_probability_set. See below for an example of their use. Further information on
the data format can be found by using the command

```{r, eval=FALSE}
help(COYU_dataset_format, package="coyu").
```

## Output

Running function COYU_all_results produces a complex output data
set. This consists of a list with one element for each probability
set. To facilitate presentation of results, there are three functions:
COYU_print_results, COYU_results_as_dataframe and plotResults.

The function COYU_print_results produces accessible tables of results
­ however at the moment these are just summaries. For this exercise
the function COYU_results_as_dataframe will be more useful. This
collates some of the more useful results in a data-frame with one row
for each candidate-by-character combination.  The function plotResults
produces plots for each character and year combination



# Examples of use

## Example 1

The first example takes a pre-existing data-set, available within the package, and shows how
to run the COYU analysis. In the code below, pathname is a suitable working directory. The

data command accesses the example data-set ­ you may wish to examine this (more
information in "Getting Started" vignette).


```{r, eval=FALSE}
library(coyu)
data(test_2_year,package="coyu")

# from example data set in package
results1<-COYU_all_results(test_2_year$trial_data,
                           test_2_year$coyu_parameters,
                           test_2_year$probability_sets)[[1]]
# note [[1]] selects the results for the first probability set

COYU_print_results(results1,        
                   test_2_year$coyu_parameters,
		   test_2_year$character_key, 
		   test_2_year$probability_set[1,])

# note test_2_year$probability_set[1,] gives the probabilities for this set

write.csv(COYU_results_as_dataframe(results1,"2_year_reject"),
	  "tester.csv")

COYU_plot_results(results1,
	          character_key=test_2_year$character_key, 
		  plot_file="MyPlots.pdf")

# results sent to a pdf file.
```

## Example 2

The second example illustrates how to use the input utility function to create input files. You
may wish to examine the objects created at each stage,

```{r, eval=FALSE}
library(coyu)
years=c(2011,2012)
characters=c(1,2,3)
varieties=c(10,11,12)

COYU_data_skeleton(years, characters, varieties)
fake_mean_data<-rbind(matrix(3,3,3), matrix(8,3,3))
fake_stddev_data<-rbind(matrix(0.1,3,3), matrix(0.8,3,3))

trial_data<-COYU_data_skeleton(years,      
			       characters,
			       varieties,
			       mean_data = fake_mean_data, 
			       stddev_data = fake_stddev_data)

params <- COYU_parameters_from_df(trial_data,c(10))

y=COYU_probability_set(reject_3_year = c(0.05,0.02,0.01),
		       reject_2_year = c(0.05,0.02,0.01), 
		       accept_2_year = c(0.1,0.05,0.02))

COYU_sanity_check(trial_data = trial_data, 
	          coyu_parameters = param)
```


# Starting from data in DUSTNT format

If data is available in DUSTNT format, then there are tools to assist
in transforming the data ready for use with the coyu package. Note
DUSTNT is software to assist with statistical analysis for DUS
testing. It is supplied by AFBI in Northern Ireland.

It is first necessary to download further code from the 
[GitHub repository](https://github.com/BiomathematicsAndStatisticsScotland/coyus).The
file "IOroutines.R" can be found in the 
[dust_stub](https://github.com/BiomathematicsAndStatisticsScotland/coyus/dust_stub)
subdirectory.

A DUSTNT file is required to link the information together for the
run. This is the input file, called COYU9.DAT by default, as described
in the instructions for the DUSTNT COYU9 and COYUS9 modules in the
DUSNT manual. This links to a UX format file. All directory references
need to be valid. The function readCoyu9File is accessed from
IORoutines.R.

The code below shows how it can be used.

```{r, eval=FALSE}
library(coyu)
source("IORoutines.R")

input_file="C:/TEMP/COYUS9.DAT" # insert own example

setwd(dirname(input_file))

data1<-readCoyu9File(basename(input_file))

results1<-COYU_all_results(data1$trial_data,
	                   data1$coyu_parameters,
			   data1$probability_sets)[[1]]

COYU_print_results(results1,
	           data1$coyu_parameters,
                   data1$character_key[1,],
                   data1$probability_sets[1,])
```
